Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-28 survey/survey.tex
imported some theseus notes

\documentclass{article}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}
\tikzset{>=latex}

@@

@@

>>>>

\documentclass{article}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usetikzlibrary{quotes}
\tikzset{>=latex}

@@

\newcommand{\kw}[1]{{\scriptsize{\textbf{\texttt{#1}}}}}
\newcommand{\ctr}[1]{{\scriptsize{\texttt{#1}}}}
\newcommand{\inline}[3]{\ctr{#1} \textasciitilde \ctr{#2}:\ctr{#3}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theseus: Types and Simple Isomorphisms}
\label{theseusI}

We begin by presenting the core of Theseus which consists of evidently
reversible functions on user-defined datatypes.  In subsequent discourse we
will use the word `map' to refer to these logically reversible
functions. Theseus has the same built-in types as~\ensuremath{\Pi^{o}} and as conventional
functional programming languages: the empty type, the unit type, and sum and
product types. All other types, including recursive types, are user-defined
using a general mechanism for type declarations:

\lstset{
        language=haskell,
 morekeywords={Bool,Nat,Tree},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
type Bool  =  (@\ctr{True}@) | (@\ctr{False}@)
type Nat   =  0 | Succ Nat
type Tree  =  Leaf Nat | Node Tree Tree
 \end{lstlisting}
%subcode source theseus.tex:347

\noindent The declarations above witness isomorphisms between the set
\kw{Bool} and the set constructed by the constants \ctr{True} and
\ctr{False}, between the set \kw{Nat} and the set inductively constructed
using {\scriptsize{0}} and \ctr{Succ}, and between the set \kw{Tree} and the
set inductively constructed using \ctr{Leaf} and \ctr{Node}.  Thus, Theseus
type definitions are similar to, say, Haskell, type definitions. Such
definitions translate directly to the underlying \ensuremath{\Pi^{o}} using iso-recursive
type definitions. For example, the above types translate as follows:
\vspace{-20pt}
\begin{small}
\begin{multicols}{3}
\[\begin{array}{rclr}
\kw{Bool}   & = &  \mu  ~x.1 + 1 \\
\kw{Nat}   & = &  \mu  ~x. 1 + x \\
\kw{Tree}   & = &  \mu  ~x. \kw{Nat}  + x * x \\
 \end{array}\]
%subcode source theseus.tex:363

\[\begin{array}{rclr}
\mathit{unfoldBool} :& \kw{Bool}  \leftrightarrow 1 + 1 &: \mathit{foldBool} \\
\mathit{unfoldNat} :& \kw{Nat}  \leftrightarrow 1 + \kw{Nat}  &: \mathit{foldNat} \\
\mathit{unfoldTree} :& \kw{Tree}  \leftrightarrow \kw{Nat}  + (\kw{Tree}  * \kw{Tree} ) &: \mathit{foldTree} \\
 \end{array}\]
%subcode source theseus.tex:368
\end{multicols}
\end{small}

%%%%%%%%%%%%
\subsection{Simple Isomorphisms}

A Theseus program has a type of the form \ctr{a} $\leftrightarrow$ \ctr{b};
one runs such a program forward by supplying a value of type \ctr{a}. If the
program terminates, we get back a value of type~\ctr{b}. Alternately we can
run the program in reverse by supplying a value of type \ctr{b}. Given the
built-in types and the facility to introduce user-defined ones, one can
already write a few small but interesting programs. Maps that are evidently
isomorphisms can be written in the familiar pattern matching style
popularized by the family of functional languages. For example:
\begin{multicols}{2}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
(@\ctr{id}@) :: Bool (@$\leftrightarrow$@) Bool
| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{False}@)
| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{True}@)

expandBool :: Bool * a (@$\leftrightarrow$@) a + a
| (@\ctr{True}@), a   (@$\leftrightarrow$@)  (@\ctr{Left}@) a
| (@\ctr{False}@), a  (@$\leftrightarrow$@)  (@\ctr{Right}@) a

expandNat :: Nat (@$\leftrightarrow$@) Nat + 1
| 0       (@$\leftrightarrow$@)  (@\ctr{Right}@) ()
| Succ n  (@$\leftrightarrow$@)  (@\ctr{Left}@) n

(@\ctr{not}@) :: Bool (@$\leftrightarrow$@) Bool
| (@\ctr{True}@)   (@$\leftrightarrow$@)  (@\ctr{False}@)
| (@\ctr{False}@)  (@$\leftrightarrow$@)  (@\ctr{True}@)

foldBool :: a + a (@$\leftrightarrow$@) Bool * a
| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{True}@), a
| (@\ctr{Right}@) a  (@$\leftrightarrow$@)  (@\ctr{False}@), a

foldNat :: Nat + 1 (@$\leftrightarrow$@) Nat
| (@\ctr{Right}@) ()  (@$\leftrightarrow$@)  0
| (@\ctr{Left}@) n    (@$\leftrightarrow$@)  Succ n
 \end{lstlisting}
%subcode source theseus.tex:410
\end{multicols}

\vspace{-25pt}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
treeUnwind :: Tree (@$\leftrightarrow$@) Tree * Tree + (Bool + Nat)
| Node t1 t2            (@$\leftrightarrow$@)  (@\ctr{Left}@)   (t1, t2)
| Leaf 0                (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{True}@))
| Leaf (Succ 0)         (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Left}@) (@\ctr{False}@))
| Leaf (Succ (Succ n))  (@$\leftrightarrow$@)  (@\ctr{Right}@)  ((@\ctr{Right}@) n)
 \end{lstlisting}
%subcode source theseus.tex:422

Unlike the situation in a conventional functional language, the
intuition here is that the two sides of each pattern clause can be
swapped to produce the inverse of the function, i.e. as we will see
repeatedly in the paper, patterns and expressions are the same
thing. Compare for example the definitions of \ctr{expandBool} and
\ctr{foldBool}. This is the only constraint that a programmer needs to
maintain. This constraint can be ensured using the following two
rules:

\begin{enumerate}
\label{rules}
\item \emph{Non-overlapping and exhaustive coverage in pattern
  clauses.} The collections of patterns in the left-hand side (LHS) of
  each clause must be a complete non-overlapping covering of the input
  type. Similarly, the collections of patterns in the right-hand side
  (RHS) of each clause must also be a complete non-overlapping
  covering of the return type. In other words, no cases should be
  omitted or duplicated.
\item \emph{Preserve typed variables across \ensuremath{\leftrightarrow}.} Each variable that
  occurs on one side of a pattern matching clause can only appear once on
  that side and must appear exactly once on the other side and with the same
  type.
\end{enumerate}

When restricted to one side of each pattern matching clause, the rules
should be familiar and intuitive. Note that, in general, the two sides
of a pattern matching clause may have different types. Other than the
requirement of non-overlapping and exhaustive coverage, there are no
rules that constrain the types and names of constants nor of
constructors. So for example, in \ctr{foldBool}, the pattern
`\ctr{Left}~\ctr{a}' maps to a pair constructor with the first
component being the constant \ctr{True}.

Examples of expressions that violate the constraints follow:

\begin{multicols}{2}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
-- Invalid: Missing LHS cases
missing_node :: Tree (@$\leftrightarrow$@) Nat
| Leaf n (@$\leftrightarrow$@) n

-- Invalid: Overlapping RHS cases
overlapping_cases :: Nat (@$\leftrightarrow$@) Nat
| 0       (@$\leftrightarrow$@)  0
| Succ n  (@$\leftrightarrow$@)  n
 \end{lstlisting}
%subcode source theseus.tex:471

\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
-- Invalid: n is dropped
drop_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
| Node t1 t2  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t1,t2)
| Leaf n      (@$\leftrightarrow$@)  (@\ctr{Right}@) 0

-- Invalid: t is used twice
dup_var :: Tree (@$\leftrightarrow$@) Tree * Tree + Nat
| Node t t  (@$\leftrightarrow$@)  (@\ctr{Left}@)  (t,t)
| Leaf n    (@$\leftrightarrow$@)  (@\ctr{Right}@) n
 \end{lstlisting}
%subcode source theseus.tex:485
\end{multicols}

\noindent It is easy to see that every primitive isomorphism of \ensuremath{\Pi^{o}} can
be expressed in Theseus. Here are a few examples.
\vspace{-10pt}
\begin{multicols}{2}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
assocL :: a+(b+c) (@$\leftrightarrow$@) (a+b)+c :: assocR
| (@\ctr{Left}@) a          (@$\,\,\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Left}@) a)
| (@\ctr{Right}@) ((@\ctr{Left}@) b)   (@$\leftrightarrow$@) (@\ctr{Left}@) ((@\ctr{Right}@) b)
| (@\ctr{Right}@) ((@\ctr{Right}@) c)  (@$\leftrightarrow$@) (@\ctr{Right}@) c
 \end{lstlisting}
%subcode source theseus.tex:499

\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
zeroe :: b + 0 (@$\leftrightarrow$@) b :: zeroi
| (@\ctr{Left}@) v (@$\leftrightarrow$@) v

swapTimes :: a * b (@$\leftrightarrow$@) b * a
| (a,b) (@$\leftrightarrow$@) (b,a)
 \end{lstlisting}
%subcode source theseus.tex:509
\end{multicols}

\label{translation}
\noindent
Every simple isomorphism of Theseus can be translated to \ensuremath{\Pi^{o}} as follows:
\begin{enumerate}
\item For every Theseus map \ctr{c : a} $\leftrightarrow$ \ctr{b}, we can
  define an intermediate type \ctr{ti} which is the sum of pairs of all the
  variables that occur in the LHS patterns. Since the LHS and RHS patterns
  contain the same type variables, \ctr{ti} is unique for a given \ctr{c} (up
  to ordering). For example, in \ctr{treeUnwind} the first clause has
  variables \lstinline{t1 : Tree} and \lstinline{t2 : Tree}, the second and
  third clauses have no variables and the fourth clause has the variable
  \lstinline{n : Nat}. The expected type \ctr{ti} is therefore
  \lstinline{Tree * Tree + (1 + (1 + Nat))}, where we insert the type
  {\scriptsize{1}} as a placeholder for the clauses that did not have type
  variables.
\item By construction \ctr{ti} is isomorphic to \ctr{a} and \ctr{b}. Hence,
  maps \ctr{c1 : a} $\leftrightarrow$ \ctr{ti} and \ctr{c2 : ti}
  $\leftrightarrow$ \ctr{b} must be expressible in \ensuremath{\Pi^{o}}. (Both the above
  isomorphisms can be constructed by systematically unfolding and
  distributing \ctr{a} and \ctr{b} until \ctr{ti} is obtained.  The exact
  details of such an expansion can be found in Sec.~3 of our previous
  paper~\cite{rc2012} and are skipped in this paper.) Once values of type
  \ctr{a} and \ctr{b} can be mapped to values of type \ctr{ti}, the required
  translation of \ctr{c} is given by \ctr{c1} $\fatsemi$ \ctr{c2}.
\end{enumerate}

%%%%%%%%%%%%
\subsection{Dealing with Numbers}

If we try to do arithmetic using the previously defined Peano-style \ensuremath{\kw{Nat} },
we have to address an issue of what to do when we encounter
\lstinline{sub1 0}. There are two obvious choices:
\begin{enumerate}
\item We can define \ctr{add1} and \ctr{sub1} of type \kw{Nat}
  $\leftrightarrow$ \kw{Nat}, such that \ctr{sub1} of {\scriptsize{0}}
  diverges. We show how these maps can be defined in Sec.~\ref{add1}.
\item An error mechanism can be added to the language such that when
  an error is raised program execution is undefined and there is no
  meaningful inverse definition.
\end{enumerate}
Another choice, similar to that made by Janus, is to use bounded integers
such that every operation is always well defined through underflows and
overflows. Here is a simple 4-bit \kw{Nat4} datatype with its corresponding
\ctr{add1} and \ctr{sub1} operations:
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool,Nat4},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
type Nat4 = Bool * Bool * Bool * Bool

add1 :: Nat4 (@$\leftrightarrow$@) Nat4 :: sub1
| (a, b, c, (@\ctr{False}@))          (@$\leftrightarrow$@)  (a, b, c, (@\ctr{True}@))
| (a, b, (@\ctr{False}@), (@\ctr{True}@))       (@$\,\leftrightarrow$@)  (a, b, (@\ctr{True}@), (@\ctr{False}@))
| (a, (@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@))    (@$\,\,\leftrightarrow$@)  (a, (@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@))
| ((@\ctr{False}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))  (@$\leftrightarrow$@)  ((@\ctr{True}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
| ((@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@), (@\ctr{True}@))   (@$\leftrightarrow$@)  ((@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@), (@\ctr{False}@))
 \end{lstlisting}
%subcode source theseus.tex:567

\noindent It is easy to see how a math library may be defined in this
way. Once an operator can be defined in Theseus it is always possible to
replace its implementation by an equivalent one that is expressible
efficiently in the underlying hardware. In this case one could compile down
to the CPU's native integer representation and operators.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Parametrized Maps}
\label{theseusII}

Now that we can express simple standalone isomorphisms, we explain how to
compose such isomorphisms to model more complex behavior. In \ensuremath{\Pi^{o}}, there
are three ways of composing isomorphisms: sequential composition, parallel
composition, and choice composition. The common idiom underlying these
composition combinators is that a reversible map can be parametrized by
another reversible map. This idea is related to ``higher-order functions''
but is more limited as we explain below.

%%%%%%%%%%%%
\subsection{Definition and Examples}

A Theseus map \ctr{f} can be parametrized by another map \ctr{g} by adding a
labeled argument~\ctr{g} of the appropriate type to \ctr{f}. In the example
below \ctr{treeUnwindf} is parametrized over some map \ctr{f} which it
applies to \ctr{n}, if the supplied tree is \lstinline{Leaf n}:
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
treeUnwindf :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
| Node t1 t2 (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
| Leaf n     (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)
 \end{lstlisting}
%subcode source theseus.tex:600

\noindent This parametrization should be thought of as a macro or a
meta-language construction. Theseus does not have high-order maps in the
formal sense. In other words, the final type of a Theseus program must be of
the form \ctr{a} $\leftrightarrow$ \ctr{b} and every occurrence of an arrow
type $\rightarrow$ must be instantiated at compile time. The notation for
instantiating the label parameter \ctr{f} in the map \ctr{fun} by the map
\ctr{g} is \inline{fun}{f}{g}. For example,
\inline{treeUnwindf}{f}{expandNat} should be thought of as shorthand for the
map:
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
_ :: Tree (@$\leftrightarrow$@) Tree * Tree + (1 + Nat)
| Node t1 t2    (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
| Leaf 0        (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Right}@) ())
| Leaf (Succ n) (@$\leftrightarrow$@) (@\ctr{Right}@) ((@\ctr{Left}@) n)
 \end{lstlisting}
%subcode source theseus.tex:618

\noindent which inlines \ctr{expandNat} within the definition of
\ctr{treeUnwindf}.

We discuss the reversibility of parametrized maps in the next section. We
first point out that the \ensuremath{\Pi^{o}} closure primitives can be expressed as
parametrized maps:
\lstset{
        language=haskell,
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
_._ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(b (@$\leftrightarrow$@) c) (@$\rightarrow$@) a (@$\leftrightarrow$@) c
| a (@$\leftrightarrow$@) g (f a)

_*_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a * c (@$\leftrightarrow$@) b * d
| (a,c) (@$\leftrightarrow$@) (f a , g c)

_+_ :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) g:(c (@$\leftrightarrow$@) d) (@$\rightarrow$@) a + c (@$\leftrightarrow$@) b + d
| (@\ctr{Left}@) a   (@$\leftrightarrow$@)  (@\ctr{Left}@) (f a)
| (@\ctr{Right}@) c  (@$\leftrightarrow$@)  (@\ctr{Right}@) (g c)
 \end{lstlisting}
%subcode source theseus.tex:637

As an additional example, parametrized maps allow us to define controlled
operations in the tradition of reversible circuits. In the definition below,
the boolean input is a control bit that determines which of the maps \ctr{th}
or \ctr{el} is applied to the second component of the pair. This conditional
is then used to define the ``controlled-not'' gate:

\lstset{
        language=haskell,
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
(@\ctr{if}@) :: th:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) el:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) Bool * a (@$\leftrightarrow$@) Bool * b
| (@\ctr{True}@), a  (@$\leftrightarrow$@) (@\ctr{Left}@) (th a)
| (@\ctr{False}@), a (@$\leftrightarrow$@) (@\ctr{Right}@) (el a)

cnot :: Bool * Bool (@$\leftrightarrow$@) Bool * Bool
| control, bit (@$\leftrightarrow$@) if ~th:(@\ctr{not}@) ~el:(@\ctr{id}@) (control, bit)
 \end{lstlisting}
%subcode source theseus.tex:653

%%%%%%%%%%%%
\subsection{Reversing Parametrized Maps}

Reversing a map of type \ctr{a} $\leftrightarrow$ \ctr{b} results in a map of
type \ctr{b} $\leftrightarrow$ \ctr{a} which is its adjoint. As discussed
before, simple maps can be reversed by simply switching the LHS and RHS for
both the type and the pattern matching clauses comprising the
body. Interestingly, one can think about the reverse of parametrized maps in
the same way -- i.e. one simply needs to switch LHS and RHS of the `\ensuremath{\leftrightarrow}'.

\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
| Node t1 t2  (@$\leftrightarrow$@) (@\ctr{Left}@) (t1, t2)
| Leaf n      (@$\leftrightarrow$@) (@\ctr{Right}@) (f n)

-- the adjoint of treeUnfold
rev_treeUnfold :: f:(Nat (@$\leftrightarrow$@) a) (@$\rightarrow$@) Tree * Tree + a (@$\leftrightarrow$@) Tree
| (@\ctr{Left}@) (t1, t2)  (@$\leftrightarrow$@) Node t1 t2
| (@\ctr{Right}@) (f n)    (@$\leftrightarrow$@) Leaf n
 \end{lstlisting}
%subcode source theseus.tex:676

\noindent Note that the application \lstinline{f n} happens on the left of
the \ensuremath{\leftrightarrow}, which is something that one does not see in conventional
functional languages. These applications in the LHS have the dual meaning of
applications in the RHS and should be understood as follows: the value that
is the actual argument of the \ctr{Right} constructor is the result of
applying~\lstinline{f n}. In the forward execution of
\lstinline{rev_treeUnfold}, when the actual value of the constructor argument
is encountered, it is passed through the reverse of \ctr{f} to get the value
\ctr{n}. In other words, the above program is exactly the same as
\lstinline{rev_treeUnfold'} below where \lstinline{rev_f} is the adjoint of
\ctr{f}:

\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
rev_treeUnfold' :: rev_f:(a (@$\leftrightarrow$@) Nat) (@$\rightarrow$@) Tree (@$\leftrightarrow$@) Tree * Tree + a
| (@\ctr{Left}@) (t1, t2) (@$\leftrightarrow$@) Node t1 t2
| (@\ctr{Right}@) a       (@$\leftrightarrow$@) Leaf (rev_f a)
 \end{lstlisting}
%subcode source theseus.tex:696

\noindent
It follows that the following programs are equivalent:
\vspace{-10pt}
\begin{multicols}{2}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
g :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
| a (@$\leftrightarrow$@) f a

g :: rev_f:(b (@$\leftrightarrow$@) a) (@$\rightarrow$@) a (@$\leftrightarrow$@) b
| rev_f b (@$\leftrightarrow$@) b
 \end{lstlisting}
%subcode source theseus.tex:710
\end{multicols}
\vspace{-15pt}
\noindent
Based on the above observation, it is easy to define the \ctr{sym} map that
constructs the adjoint of any given map:
\vspace{-5pt}
\lstset{
        language=haskell,
 morekeywords={Nat,Tree,Int,Bool},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
sym :: f:(a (@$\leftrightarrow$@) b) (@$\rightarrow$@) b (@$\leftrightarrow$@) a
| (f a) (@$\leftrightarrow$@) a
 \end{lstlisting}
%subcode source theseus.tex:722
While parametrized maps add tremendous programming convenience to Theseus,
they don't change the expressive power of the language. All programs
expressible with parametrized maps, can be expressed without them by fully
inlining the actual parameters. Thus compiling Theseus programs with
parametrized maps to \ensuremath{\Pi^{o}} simply requires first inlining the maps and then
doing the translation described in Sec.~\ref{translation}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Iteration Labels}
\label{theseusIII}
\label{add1}

The fragment of Theseus discussed up to now, modulo full recursive types, can
be expressed in pure \ensuremath{\Pi } without \ensuremath{\mathit{trace}}.  Introducing recursive
definitions in a reversible language is subtle because every iteration must
be reversible which requires, at least, that the number of iterations be
recoverable from the output. The insight we use in Theseus is that this can
be achieved elegantly by adding \emph{typed labels} as explained below. We
start with a small example:

\lstset{
        language=haskell,
 morekeywords={iter,Nat},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
parity :: Nat * Bool (@$\leftrightarrow$@) Nat * Bool
| n, b                (@$\leftrightarrow$@) iter $ n, 0, b
| iter $ Succ n, m, b (@$\leftrightarrow$@) iter $ n, Succ m, (@\ctr{not}@) b
| iter $ 0, m, b      (@$\leftrightarrow$@) m, b
  where iter :: Nat * Nat * Bool
 \end{lstlisting}
%subcode source theseus.tex:751

\noindent The map \ctr{parity} calculates whether the given natural number is
even or odd by counting to {\scriptsize{0}} and flipping the second boolean
input each time. In the definition above, the label appears twice on the LHS
and twice on the RHS. The correctness (i.e., reversibility) of the map is
guaranteed as labels obey the same restrictions as the one for simple
patterns. In particular the occurrences of the label in the LHS must
constitute a non-overlapping coverage of the label type, and similarly for
the occurrences of the label in the RHS.

The important thing to note about labels is that they may have a
different type from the LHS or RHS types of their containing map. In
this case for example, the \ctr{parity} map has \lstinline{Nat * Bool}
as its LHS and RHS type while the type of the label \kw{iter} has the
type \lstinline{Nat * Nat * Bool}. Labels temporarily give us a
different \emph{view}~\cite{Wadler:1987:VWP:41625.41653} of a value,
by changing it type. Labels act as goto statements from one side to
the other -- when a label is encountered on the RHS of a map, the
argument to the label is matched by some definition of the label
pattern on the LHS of the map, and vice versa.  Here is a trace of the
execution of \ctr{parity} as it transforms input \ctr{Succ (Succ (Succ
  0)), False} to the output \ctr{Succ (Succ (Succ 0)), True}. It takes
five pattern match and rewrite steps to complete.

\lstset{
        language=haskell,
 morekeywords={iter,Nat},
	keywordstyle=\bfseries\ttfamily,
	identifierstyle=\ttfamily,
	commentstyle=\ttfamily,
	stringstyle=\ttfamily,
	showstringspaces=false,
	basicstyle=\scriptsize\sffamily,
	tabsize=2,
	breaklines=false,
	breakatwhitespace=false,
	aboveskip={1.2\baselineskip},
        columns=fixed,
        extendedchars=true,
}
\begin{lstlisting}[escapeinside={(@}{@)}]
Succ (Succ (Succ 0)), False           (@$\longmapsto$@) iter $ Succ (Succ (Succ 0)), 0, False
iter $ Succ (Succ (Succ 0)), 0, False (@$\longmapsto$@) iter $ Succ (Succ 0), Succ 0, True
iter $ Succ (Succ 0), Succ 0, True    (@$\longmapsto$@) iter $ Succ 0, Succ (Succ 0), False
iter $ Succ 0, Succ (Succ 0), False   (@$\longmapsto$@) iter $ 0, Succ (Succ (Succ 0)), True
iter $ 0, Succ (Succ (Succ 0)), True  (@$\longmapsto$@) Succ (Succ (Succ 0)), True
 \end{lstlisting}
%subcode source theseus.tex:784

A trace operator can be expressed using labels and the definition closely
follows the expected iteration semantics of \ensuremath{\mathit{trace}} in our previous
work~\cite{James:2012:IE:2103656.2103667}.  Using \ctr{trace} one can define
\ctr{add1} on \kw{Nat} such that \ctr{sub1} of {\scriptsize{0}}
diverges.
